Towards a Complete Model of Mutation Selection Balance in Humans

### 2023-05-13 manuscript/manuscript.tex
some good stuff

The continual influx of new mutations into populations is the ultimate source
of all adaptations, but the vast majority of mutations either do not affect
fitness or are deleterious. 
Natural selection works to eliminate these deleterious mutations from the population,
thus we expect them to appear at low frequencies (Haldane?Muller?) within populations
and be less likely to fix between lineages. 
The genome itself is the product of over a billion years of evolutionary optimization,
so even though they are rare within populations, we expect that deleterious variants
should be the overwhelming majority of segregating variation in functional regions,
and be responsible for a significant portion of phenotypic variation and disease (XX cites XX),

As with most selection in the genome, selection on deleterious variation
process perturbs the allele frequencies of neighboring linked variation. This
reduces genetic variability around conserved segments and creates large-scale
spatial patterns in genetic diversity. These patterns are mediated by the
deleterious mutation rate, the strength of selection in conserved segments, and
the spatial distribution of recombination rates and conserved segments. Since
genome-wide recombination maps and putatively conserved segments are available
for many species, researchers have combined these with population genetic
theory to estimate the mutation rate and selection coefficients of deleterious
mutations, as well as the overall reduction in genetic variability due to the
negative selection process.

@@

When mutations are weakly deleterious, their allele frequencies are strongly
influenced by stochastic perturbations and their frequency trajectories are no
longer well-approximated by deterministic models. When genetic drift is the
only source of randomness in allele frequency change, the relevant scale of
stochastic perturbations is determined by the drift-effective population size
$N_d$ (e.g. due to only non-selective process). For reasons that will soon be
clear, we consider a general expression of the drift-effective population size
that allows for extra non-heritable variability in offspring number $V_k$,

\begin{align}
    N_d = \frac{2N}{V_k + 2}
\end{align}

where if $V_k = 2$, this reduces to the classic Wright--Fisher model
\parencite{Wright1938-tv}.

@@

To further complicate matters, both strong and weakly deleterious mutations
create \emph{heritable} fitness variation. In turn, the presence of heritable
fitness variation generates another source of randomness in allele frequency
change known as genetic \emph{draft} \parencite{Neher2013-dz}. Genetic draft
occurs when alleles become randomly associated with different fitness
backgrounds each generation, which stochastically perturbs their trajectories.
Like genetic drift, perturbations due to draft are directionless but overall
increase the variance in allele frequency change. Unlike drift, the stochastic
perturbations each generation due to draft are autocorrelated as long as
segregation and recombination do not decouple an allele from its fitness
background and the selective environment does not change
\parencite{Robertson1961-ho,Santiago1995-hx,Buffalo2019-qs}. When selection is
strong,

When selection is
weak and fitness is polygenic, the stochasticity introduced by draft is
well-approximated by rescaling the drift-effective population size

that exist when there are heritable differences among haplotypes.
Mathematical models of genetic draft

Gillespie2000-mh

,Robertson1961-ho,Santiago1995-hx,Santiago1998-bs,Barton2000-zg

\subsection*{DFE Estimates}
 - Table s5 %https://oup.silverchair-cdn.com/oup/backfile/Content_public/Journal/genetics/206/1/10.1534_genetics.116.197145/8/files1.pdf?Expires=1686357613&Signature=cp9qoI2ChCb6GFZ74HI4eWcA~MQ4sHlWyeI3Zp49-5xEORzMfYlSRxzVvZlo8ZoGWDi0G3PnL00ZplM-~or4YIVm2uhkmqjf61AmC7RdZ1QTak4j0kFMQ0wAbM9pusiV3gIepTQHq2HKRKwgS~ypmv~0oqO5dDxaq9nhOWIBBmLAA49oHWuxUtnRe3sEYATwCwjIFEomhgJpg7vKGCF~87GdCV4JETW4PN2VvD4hsu2AVRd8yA~vUJvsXfhWwcElDeY4fsgpprl1lC3XGjJCnqdwbzRGPV5LNx5XEKLptcZtLeLcOmFJCvmVLe0bewJ7Vta0TSQM7T9AuklompjPPw__&Key-Pair-Id=APKAIE5G5CRDK6RD3PGA

@@

We first assessed the B' maps against the observed average reduction across
$XXX$ forward-simulations, $\bar{B}_\text{sim}(x)$. The mean-squared error has
two sources, (1) bias from a breakdown in our theory $B(x)$ for the average
reduction, and (2) variance from the simulations. We show that for most of the
parameter space, the measured mean-squared error between simulation averages
and our B' maps is at the theoretic minimum. This minimum variance is entirely
from the coalescent noise when estimating $\bar{B}\text{sim}(x)$, which

@@

In their 2016 paper, Santiago and Caballero derive a general equation for the
fitness variance that accounts for the reduction due to the fixation of weakly
deleterious alleles (i.e. the rate of Muller's ratchet). There derivation for
the variance uses a heuristic approach based on Fisher's Fundamental Theorem of
Natural Selection (equation 2, \cite{Santiago2016-mu}, equation 15A
%`'\cite{Garcia-Dorado2007-jj}). 
A more formal derivation follows from \textcite{Higgs1995-xc} 
(though they assume a haploid asexual system, the end
result is the same up to a factor of two). In their work, they derive equations
for the change per generation for the moments and cross-moments of the fitness
distribution under multiplicative selection in a haploid model. They find, the
haploid rate of the ratchet is $r = u - s v_n$ (equation 13.3,
\cite{Higgs1995-xc}), where we use lowercase to distinguish between haploid and
diploid models. Here, $v_n$ is the variance in the \emph{number} of deleterious
mutations; as long as selection is not too strong, the haploid additive
variance is $v_{A} \approx s^2 v_n$. Substituting this for $v_n$ and
rearranging, the haploid fitness variance is related to the $v_A = us - rs$,
which is identical to Santiago and Caballero's (\citeyear{Santiago2016-mu})
equation 2, noting that the waiting time between fixations (i.e. ratchet
clicks) is $T = \nicefrac{1}{r}$. Intuitively, the fitness variance $Us$ is
reduced by the rate at which deleterious mutations fix and remove fitness
variation from the population.

@@

%\printbibliography
\end{document}

>>>>

The continual influx of new mutations into populations is the ultimate source
of all adaptations, but the vast majority of mutations either do not affect
fitness or are deleterious. Negative selection works to eliminate such
deleterious mutations from the population, which is essential to maintaining
the integrity of information in the genome to maintain a lineage. Mutations
under negative selection are kept at low frequencies (Haldane?), contributing
to human traits (XXX) and diseases. As with most selection in the genome, this
process perturbs the allele frequencies of neighboring linked variation. This
reduces genetic variability around conserved segments and creates large-scale
spatial patterns in genetic diversity. These patterns are mediated by the
deleterious mutation rate, the strength of selection in conserved segments, and
the spatial distribution of recombination rates and conserved segments. Since
genome-wide recombination maps and putatively conserved segments are available
for many species, researchers have combined these with population genetic
theory to estimate the mutation rate and selection coefficients of deleterious
mutations, as well as the overall reduction in genetic variability due to the
negative selection process.

@@

When mutations are weakly deleterious, their allele frequencies are strongly
influenced by stochastic perturbations and their frequency trajectories are no
longer well-approximated by deterministic models. When genetic drift is the
only source of randomness in allele frequency change, the relevant scale of
stochastic perturbations is determined by the drift-effective population size
$N_d$ (e.g. due to only non-selective process). For reasons that will soon be
clear, we consider a general expression of the drift-effective population size
that allows for extra non-heritable variability in offspring number $V^2$,
%
\begin{align}
    \label{eq:simple_Ne}
    N_d = \frac{2N}{V^2 + 2}
\end{align}

where if $V^2 = 2$, this reduces to the classic Wright--Fisher model
\parencite{Wright1938-tv}.

@@

To further complicate matters, both strong and weakly deleterious mutations
create \emph{heritable} fitness variation. In turn, the presence of heritable
fitness variation generates another source of randomness in allele frequency
change known as genetic \emph{draft} \parencite{Neher2013-dz}. Genetic draft
occurs when alleles become randomly associated with different fitness
backgrounds each generation, which stochastically perturbs their trajectories.
Like genetic drift, perturbations due to draft are directionless but overall
increase the variance in allele frequency change. Unlike drift, stochastic
perturbations caused by draft are autocorrelated across generations as long an
allele does not recombine or segregate off its fitness background and the
selective environment does not change
\parencite{Robertson1961-ho,Santiago1995-hx,Buffalo2019-qs}. When selection is
strong, draft can generate multiple-merger coalescences that create
discontinuities in allele frequency trajectories, making diffusion
approximations ill-suited \parencite{Gillespie2000-mh,Der2011-it,Neher2013-dz}.
However, when selection is weak and polygenic, draft can be approximated as an
increase in the rate of drift. The key insight is that the draft rescaling
factor is like extra variability in offspring number, but because of
autocorrelation due to linkage and similar environments, the heritable fitness
variation $C^2$ (on a relative fitness scale, i.e. the squared coefficient of
variance) becomes inflated by a factor $Q^2$. Intuitively, the product $V^2
Q^2$ represents the expected total variance a neutral mutation experiences over
its lifetime in a system with draft. This \emph{heritable} fitness variance
that neutral allele experiences can then be included in Equation
\eqref{eq:simple_Ne} (assuming the usual Wright--Fisher non-heritable variance,
i.e. $V^2=2$),

\begin{align}
    \label{eq:main_Ne}
    N_e = \frac{N}{C^2 Q^2 + 1}.
\end{align}

(c.f. \cite{Robertson1961-ho,Santiago1995-hx}; see XXX for a proof). Note that
because the variance under drift increases each generation due to the
accumulation of drift perturbations and draft autocorrelations, the inflation
term $Q^2$ has a temporal dimension. For notational efficiency, we assume $Q^2$
refers to the asymptotic $Q_t^2$ as $t \to \infty$ and $N_e$ is the
\emph{asymptotic} draft effective population size. In general, however,
statistics such as heterozygosity depend on the cumulative autocorrelation
factor \emph{up to some time} $t$, $Q_t^2$ \parencite[p.
2111]{Santiago1998-bs}. This reflects the fact that effective population size
experienced by a mutation newly associated with a fitness background is larger
than the asymptotic effective population size, since not much autocorrelation
has accumulated in the stochastic perturbations caused by drift yet. From a
backwards-time perspective, this reflects the fact that pairwise diversity
(which is approximately heterozygosity in most cases) depends on the pairwise
coalescence rate per generation, which will is not constant under draft. The
benefit of using a forward-time model of draft is that it the inflation factor
is invariant with respect to the particular fitness background the neutral
allele becomes stochastically associated with. By contrast, the difficulty with
backwards-time models is that the coalescence rates experienced by a lineage
are not invariant to which lineage was sampled.

Equation \eqref{eq:main_Ne}

Shortly, we will show that Robertson (1961) and Santiago and Caballero's (1995,
1998, 2016) approach

rescaling the drift-effective population size by an appropriate factor
\parencite{Robertson1961-ho,Santiago1995-hx,Santiago1998-bs,Barton2000-zg}.
While a simple rescaling cannot elucidate details about how selection alters
the genealogy, it does allow for the calculation of key quantities like the
pairwise coalescence time (though see \textcite{Santiago1998-bs}, p. 2111).

\subsection*{DFE Estimates}
% - Table s5 https://oup.silverchair-cdn.com/oup/backfile/Content_public/Journal/genetics/206/1/10.1534_genetics.116.197145/8/files1.pdf?Expires=1686357613&Signature=cp9qoI2ChCb6GFZ74HI4eWcA~MQ4sHlWyeI3Zp49-5xEORzMfYlSRxzVvZlo8ZoGWDi0G3PnL00ZplM-~or4YIVm2uhkmqjf61AmC7RdZ1QTak4j0kFMQ0wAbM9pusiV3gIepTQHq2HKRKwgS~ypmv~0oqO5dDxaq9nhOWIBBmLAA49oHWuxUtnRe3sEYATwCwjIFEomhgJpg7vKGCF~87GdCV4JETW4PN2VvD4hsu2AVRd8yA~vUJvsXfhWwcElDeY4fsgpprl1lC3XGjJCnqdwbzRGPV5LNx5XEKLptcZtLeLcOmFJCvmVLe0bewJ7Vta0TSQM7T9AuklompjPPw__&Key-Pair-Id=APKAIE5G5CRDK6RD3PGA

@@

We first assessed the B' maps against the observed average reduction across
$XXX$ forward-simulations, $\bar{B}_\text{sim}(x)$. The mean-squared error has
two sources, (1) bias from a breakdown in our theory $B(x)$ for the average
reduction, and (2) variance from the simulations. We show that for most of the
parameter space, the measured mean-squared error between simulation averages
and our B' maps is at the theoretic minimum. This minimum variance is entirely
from the coalescent noise when estimating $\bar{B}\text{sim}(x), which

@@

In their 2016 paper, Santiago and Caballero derive a general equation for the
fitness variance that accounts for the reduction due to the fixation of weakly
deleterious alleles (i.e. the rate of Muller's ratchet). There derivation for
the variance uses a heuristic approach based on Fisher's Fundamental Theorem of
Natural Selection (equation 2, \cite{Santiago2016-mu}, equation 15A
\cite{Garcia-Dorado2007-jj}). A more formal derivation follows from
\textcite{Higgs1995-xc} (though they assume a haploid asexual system, the end
result is the same up to a factor of two). In their work, they derive equations
for the change per generation for the moments and cross-moments of the fitness
distribution under multiplicative selection in a haploid model. They find, the
haploid rate of the ratchet is $r = u - s v_n$ (equation 13.3,
\cite{Higgs1995-xc}), where we use lowercase to distinguish between haploid and
diploid models. Here, $v_n$ is the variance in the \emph{number} of deleterious
mutations; as long as selection is not too strong, the haploid additive
variance is $v_{A} \approx s^2 v_n$. Substituting this for $v_n$ and
rearranging, the haploid fitness variance is related to the $v_A = us - rs$,
which is identical to Santiago and Caballero's (\citeyear{Santiago2016-mu})
equation 2, noting that the waiting time between fixations (i.e. ratchet
clicks) is $T = \nicefrac{1}{r}$. Intuitively, the fitness variance $Us$ is
reduced by the rate at which deleterious mutations fix and remove fitness
variation from the population.

@@

\printbibliography

\end{document}
